GEE models

### 2017-07-19 doc/submission/reviewer-response.md

    - *"Latent classes. Although this is a very attractive method, based on ESM
    figure 4, there are 3 parallel declining group trajectories that is mostly
    affected by the baseline values. Groups based on baseline (last follow-up)
    tertiles would be easier to interpret."* ... *"The PLS-DA analysis needs
    further description. I would expect variables with the highest loadings to
    be the variables in the unadjusted GEE analysis. I am probably wrong given
    the results."* ... *"Do we really expect good discrimination? The
    associations are week to nul between NEFAs and the outcomes. I do not really
    see the point the extensive supplementary analysis on PLS-DA. Furthermore
    dysglycemia is an outcome that is not directly related to beta-cell function
    only. Dysglycemia requires both insulin resistance and beta-cell
    dysfunction."*

        Given this reviewer's previous comments and suggestions, we have decided
        to: focus on the GEE analysis in the main paper; remove the LCMM
        analysis; perform the PLS analysis on the baseline (cross-sectional)
        data using the outcomes as continuous variables; and, briefly describe
        the findings in the Results section. To keep the ESM material
        streamlined, we have contained the PLS results and other additional
        analyses conducted within the code (accessible via the DOI) for the
        interested reader. We believe this simplifies the results and primary
        findings of this paper without sacrificing a loss in the potential use
        these additional analyses may provide for the interested researcher.

    - *"The authors mention in the Discussion the external validity is mostly
    limited to white females. I suggest to run a sensitivity analysis on this
    group."*

        We agree with the reviewer's comment and ran stratified models by either
        sex or ethnicity. In both cases, all significant associations from the
        non-stratified analyses were attenuated, likely due to the reduction in
        sample size in the stratified models. Stratifying analyses has the
        limitation of substantially reducing sample size and is the reason why
        interaction testing is prefered. We ran interaction tests by sex and
        ethnicity and found no significant interactive associations. While
        these results internally suggest no differences between sex or ethnicity,
        our cohort was not designed to answer these research questions and the
        results presented in the paper may still not be generalizable to other
        populations. {{yea?}}

@@

8. *"DISCUSSION"*
    - *"I would like to see more on the DAGs that underlies the selected
    variables in the models. That would either confirm the way the analysis is
    currently performed, or would show that some of the adjustments are
    over-adjustment."*

        As requested, we have included the DAGs underlying the hypothesized
        associations in the ESM.

@@

    - *"In the time-varying analysis some covariates are measured after NEFAs,
    thus adjustment for them could be over-adjustment just because they are
    measured closer to the outcome of interest."*

        In order to confirm the impact the comment made by the reviewer has on
        the reported associations, we ran sensitivity analyses where we had all
        covariates as time-independent, to match the NEFA variables in the GEE
        model. We found that there were negligible differences in the results
        between models with covariates as either time-dependent and
        time-independent and models with all covariates as time-independent.

### 2017-07-20 doc/submission/reviewer-response.md

# Responses to reviewers

## Referee: 1

- *"First, I would like to congratulate you for the exhaustive and complete
research work that has been done for this paper. I think the statistical study
is very successful and useful for researchers who have accumulated a large
database and need a method with the ability to handle a large number of
variables and their interactions. I really liked read in the discussion the
negative points of the study, with which I fully agree, and not an attempt to
hide the limitations of any scientific work. In retrospect, perhaps it will miss
data that clarify the pre-diabetic state, or not, of the patients (HbA1c or
C-peptide). I think it would be good for the study know if some patients were
taking any medication that could affect the results. Although it is a
complicated scientific work to understand for researchers who do not usually use
this kind of statistical studies, I believe that is described in a very
appropriate and clear manner."*

    We thank the reviewer for these comments. {{not sure about comment on
    pre-diabetes}}. In the PROMISE cohort, we collected medication data from
    all participants. While most participants took some form of medication over
    the 6 years, during the times of data collection, few ({{get number}}) were
    on medications. However, for those who were taking medications that could
    influence the results, additional analyses revealed that medication use did
    not substantially influence the predictor or outcome measures, nor did 
    medication use substantially contribute to GEE model fit compared to models 
    without it (see QIC table).

### 2017-07-20 doc/manuscript.Rmd
made edits to abstract

# Abstract
<!--243 / 250 words-->

@@

Methods: In the PROMISE cohort, 477 participants had serum
NEFA measured at the baseline visit and completed an OGTT at 3 time points over
6 years. Outcome variables were calculated using the OGTT values. IS was
assessed using HOMA2-%S and the Matsuda index, while beta-cell function was
assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were
used, adjusting for time, waist, sex, ethnicity, baseline age, alanine
aminotransferase, and family history of diabetes. NEFA were analyzed as both
concentrations (nmol/mL) and proportions (mol%) of the total fraction.

Results: IS and beta-cell function declined by 14-27% while BMI did not
significantly change over 6 years of follow-up. Total NEFA, 16:0, 18:1n-9, and
18:2n-6 as concentrations had significant negative associations with IGI/IR and
ISSI-2, predicting between 4-8% lower IGI/IR and ISSI-2 over the 6 years.
However, individual NEFA modeled as mol% showed no associations with these
outcomes. Total and individual NEFA were not associated with IS measures.

>>>>

# Abstract

@@

Methods: In the PROMISE cohort, 477 participants had serum NEFA (mol% and
concentration) measured at the baseline visit and completed an OGTT at 3 time
points over 6 years. Outcome variables were calculated using the OGTT values. IS
was assessed using HOMA2-%S and the Matsuda index, while beta-cell function was 
assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin 
Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were 
used, adjusting for time, waist, sex, ethnicity, baseline age, alanine 
aminotransferase, and physical activity. NEFA were analyzed as both 
concentrations (nmol/mL) and proportions (mol%) of the total fraction.

Results: Participants (73% female, 70% with European-ancestry) had their IS and
beta-cell function decline by 14-27% over 6 years of follow-up. While in
unadjusted models there were several NEFA (e.g. 18:1n-7, 20:1n-9, 22:4n-6) that
had negative associations with insulin sensitivity, nearly all of these
associations were attenuated in fully adjusted models. In adjusted models, total
NEFA, 16:0, 18:1n-9, and 18:2n-6 (as concentrations) had significant negative
associations with IGI/IR and ISSI-2, while only 20:5n-3 (as mol%) had positive
associations with HOMA2-%S.
